---
title: "session07 - homework"
format: html
editor: visual
date: 2025-07-17
author: Timo Roettger
execute:
  error: false
  warning: false
  message: false
  cache: false
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, lme4, brms, bayestestR, tidybayes, gridExtra)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

# options to increase efficiency of brms models (optional)
options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

### Introduction to the problem

We will be working with Quantese again, see `01_exercise.qmd` and the other data resources for a refresher.

```{r}
#| label: load-quantese

quantese <- read_csv("../../data/quantese_naming_sim.csv") 
# add log-transformed reaction time
quantese <- quantese |> 
  mutate(log_rt = log(rt))

quantese

```

In an earlier homework, we have assessed the following hypothesis:

*"We hypothesize that log-transformed reaction times are affected by lexical frequency"*

We tried to run the following model that included an appropriate random effect structure but it did not converge with `lme4`:

```{r}
#| label: lmer_quantese

# formulate the corresponding LMER
quant_lmer <- lmer(log_rt ~ log_frequency +
                     (1 + log_frequency | PartID) +
                     (1 | ItemID), 
                    data = quantese)
# your receive a warning: "boundary (singular) fit: see help('isSingular')"

```

*So you consider singular fit to be non-conversion? Jo told us that these were two different things.*

Now we will re-run this model within the Bayesian framework:

### (A) Specify priors

Check which priors this model needs and specify weakly informative priors.

Remember that weakly informative priors are priors that do not bias the posteriors in any direction, but reasonably constrain the parameter space, i.e. only reasonable values are assumed.

For priors of class `Intercept` and `b`, constrain the parameter space with a wide normal distribution.

```{r}
#| label: get_prior

# YOUR CODE HERE
model_formula <- bf(log_rt ~ log_frequency)
get_prior(model_formula, quantese)
# So I need intercept and b, and possibly I should have sigma too but idk.

# Plot the log_rt values in the data bc I have no idea what's a reasonable range
ggplot(data = quantese,
       aes(x = log_rt)) +
  geom_density()
# To be as vague as possible, try and cover much more than this range, starting in
# the middle (not the peak) for possible intercepts

# specify weakly informative priors
my_priors <- c(prior(normal(6.75, 2),
                     class = 'Intercept'),
               # Just picked half-ish the range I see in the data for spread,
               # bc that's real broad
               prior(normal(0, 1),
                     class = 'b'))

```

We can also now get used to specifying priors for classes `sd` and `cor`. Priors of class `sd` are priors for the variance associated with our random effects. You see here 2 `sd` priors for group `ItemID`. The first one is just a generic prior for all `ItemID` coefficients and the second one with the `coef = Intercept` stands for the random intercept variation between items. Similarly for group `PartID`, we have 3 priors. The first one is again just a generic prior for all `PartID` coefficients. The second one with the `coef = Intercept` stands for the random intercept variation between participants. And the last one with the `coef = log_frequency` stands for the random slope variation between participants for the predictor `log_frequency`. How do we specify these priors. We have usually no idea what to expect from these parameters, so we here also specify very diffuse priors centered on zero. We pick a cauchy distribution (remember? super squashed normal distributions) because they have thick tails. Let's make them super diffuse and pick a `sd = 3` for both items and participants.

```{r}
#| label: add_sd_priors

# add weakly informative priors for sd
my_priors <- c(my_priors, 
               prior(cauchy(0, 3), class = 'sd', group = 'ItemID'),
               prior(cauchy(0, 3), class = 'sd', group = 'PartID'))

```

And finally, if your model as random slopes, they have by default a correlation parameter for intercept and slope. We are not getting into the math here and will follow advice from smart people. We use Lewandowski-Kurowicka-Joe priors (shape = 1): `lkj(1)`. Specifying this prior for random slope models actually helps a lot with the fitting procedure.

```{r}
#| label: add_cor_priors

# add weakly informative priors for cor
my_priors <- c(my_priors, 
               prior(lkj(1), class = 'cor'))

```

### (B) Run model

Now run the model with 4 chains, 1000 warm-up and 1000 post-warm-up samples per chain. Check that chains are reasonably mixed and that sampling yielded a sufficient effective sample size. If not address the issues appropriately (check slides again).

```{r}
#| label: model_quant

# if the model takes way too long to sample (more than 5-10 mins), use the following subset
# quantese_sub <- quantese |> 
#   filter(PartID %in% c("01","02","03","04",
#                        "05","06","07","08"))

# YOUR CODE HERE
quant_brm <- brm(log_rt ~ log_frequency +
                   (1 + log_frequency | ItemID) +
                   (1 + log_frequency | PartID),
                 # Should it still be gaussian?
                 family = gaussian(),
                 prior = my_priors,
                 data = quantese,
                 chains = 4,
                 iter = 2000, warmup = 1000)
# Should I be adding the rest of the specifications in the slides?

mcmc_plot(quant_brm, type = 'trace')
mcmc_dens_overlay(quant_brm, pars = 'b_Intercept')

```

Wow, so many caterpillars! They're too small to see super clearly, but I think they mostly look fine. The ones for `itemID_Intercept` and `ItemID_log_frequency` look a little less mixed.

Ohhhh but the density overlay does not look so good, even just for `b_Intercept`. I'll try increasing the warmup period.

```{r}
#| label: model_adjusted
quant_brm2 <- brm(log_rt ~ log_frequency +
                   (1 + log_frequency | ItemID) +
                   (1 + log_frequency | PartID),
                 family = gaussian(),
                 prior = my_priors,
                 data = quantese,
                 # MCMC specs
                 chains = 4,
                 iter = 6000, warmup = 4000)
# The warning for divergences is smaller this time (10%), but that's still > 0.

mcmc_plot(quant_brm2, type = 'trace')
mcmc_dens_overlay(quant_brm2, pars = 'b_Intercept')
# Plots look significantly better, but still I should probably try increasing adapt_delta.
# And I'll add the rest of the mcmc specs from the slides.

quant_brm3 <- brm(log_rt ~ log_frequency +
                   (1 + log_frequency | ItemID) +
                   (1 + log_frequency | PartID),
                 family = gaussian(),
                 prior = my_priors,
                 data = quantese,
                 # MCMC specs
                 chains = 4,
                 iter = 6000, warmup = 4000,
                 thin = 1, seed = 666,
                 control = list(adapt_delta = 0.99))
# This model took more than 10 minutes trying to sample! Had to abort.

# Doing what was advised above, then running that version.
quantese_sub <- quantese |> 
   filter(PartID %in% c("01","02","03","04",
                        "05","06","07","08"))

quant_brm4 <- brm(log_rt ~ log_frequency +
                   (1 + log_frequency | ItemID) +
                   (1 + log_frequency | PartID),
                 family = gaussian(),
                 file = '../../models/07_homework.RDS',
                 prior = my_priors,
                 data = quantese_sub,
                 # MCMC specs
                 chains = 4,
                 iter = 6000, warmup = 4000,
                 thin = 1, seed = 666,
                 control = list(adapt_delta = 0.99))

# Nice and quick this time!

mcmc_plot(quant_brm4, type = 'trace')
mcmc_dens_overlay(quant_brm4, pars = 'b_Intercept')

# Finally, the chains look excellent and the model does not complain of divergences, hurrah
# Checking sample size

quant_brm4

```

Bulk_ESS could be larger for the ItemID values (Intercept sd = 858, slope sd = 441), and Rhat = 1.01 for the latter. So it's borderline, but the rules of thumb are satisfied.

### (C) Plot posterior predictions

Find a way to extract posterior samples from the model for the `Intercept` and the `log_frequency` coefficient and plot the predicted posterior mean linear model alongside its 95% Credible Intervals. There are many different ways to get there.

```{r}
#| label: plot_predictions

# YOUR CODE HERE
get_variables(quant_brm4)

# I don't understand the difference between the variables `Intercept` and `b_Intercept`.

post_samples_quant <- quant_brm4 |>
  spread_draws(Intercept, b_log_frequency)

post_means_quant <- post_samples_quant  |>
  summarise(intercept_mean = mean(Intercept),
            slope_mean = mean(b_log_frequency))

post_intervals_quant <- post_samples_quant |>
  median_qi(Intercept, b_log_frequency)

# Plotting predicted posterior mean linear model
# Okay, I'm so confused trying to plot it as an actual line with predicted means
# And uncertainty for the intercept AND slope.

# For now I'll just plot them separately. Added patchwork library for this.
intercept_plot <- ggplot() +
  geom_point(data = post_means_quant,
                         aes(x = 1,
                             y = intercept_mean)) +
  geom_errorbar(data = post_intervals_quant,
                 aes(x = 1,
                     ymin = Intercept.lower,
                     ymax = Intercept.upper)) +
  scale_x_continuous(labels = NULL) + # The row number is meaningless
  scale_y_continuous(limits = c(5, 8)) + # This covers the range of the sample means
  labs(x = '\nintercept',
       y = 'predicted mean intercept for log reaction time\n') +
  theme_minimal()

slope_plot <- ggplot() +
  geom_point(data = post_means_quant,
                         aes(x = 1,
                             y = slope_mean)) +
  geom_errorbar(data = post_intervals_quant,
                 aes(x = 1,
                     ymin = b_log_frequency.lower,
                     ymax = b_log_frequency.upper)) +
  scale_x_continuous(labels = NULL) +
  scale_y_continuous(limits = c(-0.15, 0)) + # This covers the range of the sample slopes
  labs(x = '\nslope',
       y = 'predcted mean slope for log reaction time\n') +
  theme_minimal()

grid.arrange(intercept_plot, slope_plot,
             ncol = 2,
             top = 'Predicted posterior mean values for intercept and mean\nin quantese data subset. Error bars = 95% CI\n',
             bottom = 'parameter')
```

Sorry, I know this is not really the linear model plot you asked for and is also kind of horrible code, but it's the best I had time to achieve. I look forward to seeing the solutions.

# Useful links

<https://cloud.r-project.org/web/packages/tidybayes/vignettes/tidy-brms.html#point-summaries-and-intervals>

<https://r-charts.com/ggplot2/combining-plots/>

<https://stackoverflow.com/questions/35090883/remove-all-of-x-axis-labels-in-ggplot#comment117163007_35090981>
