---
title: "session11 - homework"
format: html
editor: visual
date: 2025-07-17
author: Timo Roettger
execute:
  error: false
  warning: false
  message: false
  cache: true
bibliography: ../resources/bibliography.bib
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, tidybayes, brms, ggdist, bayestestR, modelr)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

# options to increase efficiency of brms models (optional)
options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

# Introduction

We will be analyzing the data from @roettger2014assessing.

```{r}
#| label: preprocessing

# Load our datasets
incNeutr <- read_csv("../../data/E1_production.csv") |> 
  # we will focus only on the relevant columns
  select(subject, voicing, item_pair, usable, vowel_dur) |> 
  # and only on rows the authors deemed usable
  filter(usable == "usable") |> 
  select(!usable)

```

The variables:

`subject`: unique identifier for participants (n = 16)

`voicing`: underlying voicing status of stop (`voiced` vs. `voiceless`)

`item_pair`: unique identifier for pseudo word minimal pairs (n = 24)

`vowel_dur`: vowel duration in ms

In this study, we recorded people producing words such as Rad and Rat in German and were interested to see whether these words phonetically differ. The reason why this is interesting is that, according to German phonology, these words *should* be identical, i.e. he voicing distinction between /d/ and /t/ should have been *neutralized.* However, previous literature indicates that this neutralization is phonetically *incomplete.* For example, the vowel preceding a (de)voiced stop such /d/ is slightly longer preceding a voiceless stop such as /t/.

The researchers hypothesize that the vowel duration is meaningfully longer preceding voiced stops than preceding voiceless stops. They model this relationship using a linear mixed effects model that accounts for both speaker and word variation as well as by-speaker and by-word variation for the critical predictor voicing.

```{r}
#| label: model_formula

# store model formula
mdl_formula = bf(vowel_dur ~ voicing +
                   # random intercepts and slopes for subject and item_pair
                   (1 + voicing | subject) +
                   (1 + voicing | item_pair))

```

```{r}
#| label: spec-priors

# some weakly informative priors 
my_priors <- c(prior(normal(150,50), class = 'Intercept'),
               # weakly informative prior for the voicing difference (very diffuse)
               prior(normal(0, 30), class = 'b', coef = 'voicingvoiceless'),
               # weakly informative prior for all sources of variance (very diffuse)
               prior(cauchy(0,100), class = 'sd'),
               # weakly informative prior for residual variance (very diffuse)
               prior(cauchy(0,100), class = 'sigma'),
               # correlations
               prior(lkj(1), class = 'cor'))

```

```{r}
#| label: run-model

# run model
xmdl <- brm(mdl_formula,
            family = gaussian(),
            prior = my_priors,
            file = "../../models/in_sesoi",
            data = incNeutr,
            seed = 42)

# I also store the prior-only model 
xmdl_prior <- brm(mdl_formula,
            family = gaussian(),
            prior = my_priors,
            sample_prior = "only",
            file = "../../models/in_sesoi_priors",
            data = incNeutr,
            seed = 42)

```

# Exercises

## (a) Plot the posterior distribution

Explore the posterior distribution for the difference between voiced and voiceless stops. Interpret it.

```{r}

# YOUR ANSWERS HERE
xmdl |>
  spread_draws(b_voicingvoiceless) |>
  ggplot(aes(x = b_voicingvoiceless)) +
  stat_halfeye() +
  geom_vline(xintercept = 0, linetype = 2) +
  labs(x = '\nPosterior estimates for mean difference between voiced and voiceless conditions')
  theme_minimal()
```

## (b) Define a SESOI

Define a SESOI and justify it. Can you find an applied justification for a SESOI. Or a theoretical justification? Are there justifications grounded in measurement? What about justifications based on previous research? Define it and formalize it as a Region of Practical Equivalence as we have done before. Use the `bayesfactor_rope()` function to calculate the Bayes factor for and against the posteriors being inside the ROPE.

```{r}

# YOUR ANSWER HERE
rope_narrow <- c(-5, 5)
rope_wide <- c(-15, 15)
```

Given my lack of field knowledge, I base my choices on previous research, specifically @nicenboim2018 (p. 50). The narrow SESOI of 5ms is based on evidence that studies with greater precision consistently find duration differences of greater than 5ms. Meanwhile, the wide SESOI of 15ms reflects the largest differences found in this meta-study. I would like to include measurement-based justification, but do not understand the properties of measurement used here well.

```{r}
# Bayes factor for narrow ROPE
BF_narrow <- bayesfactor_rope(posterior = xmdl,
                              prior = xmdl_prior,
                              null = rope_narrow,
                              seed = 42,
                              parameter = "b_voicingvoiceless")

BF_wide <- bayesfactor_rope(posterior = xmdl,
                              prior = xmdl_prior,
                              null = rope_wide,
                              seed = 42,
                              parameter = "b_voicingvoiceless")

BF_narrow # ~2.69
BF_wide # <0.01
```

Ooooh, interesting! For a SESOI of 5ms, the BF is 2.69. So after seeing the data, our posterior belief that the difference between the vowel durations for underlying voiced and voiceless conditions is outside the ROPE is 2.69 times more probable than our prior belief before seeing the data. This is considered anecdotal evidence in favor of the alternative hypothesis based on @lee2014. Meanwhile, examining a SESOI of 15ms, the BF is \< 0.01, indicating extreme evidence in favor of the null that the difference falls within the ROPE. Thus, we can't make broad conclusions about the presence of an incomplete neutralisation effect based on this data, but if it does exist, it much more likely to be above 5 than above 15 ms.

## (c) Check robustness of your analysis

How robust is your conclusion against different prior choices?

```{r}

# YOUR ANSWER HERE
# Narrower distribution for prior difference between conditions
my_priors_10 <- c(prior(normal(150,50), class = 'Intercept'),
               # weakly informative prior for the voicing difference (very diffuse)
               prior(normal(0, 10), class = 'b', coef = 'voicingvoiceless'),
               # weakly informative prior for all sources of variance (very diffuse)
               prior(cauchy(0,100), class = 'sd'),
               # weakly informative prior for residual variance (very diffuse)
               prior(cauchy(0,100), class = 'sigma'),
               # correlations
               prior(lkj(1), class = 'cor'))

# Medium diffuse prior differences
my_priors_20 <- c(prior(normal(150,50), class = 'Intercept'),
               # weakly informative prior for the voicing difference (very diffuse)
               prior(normal(0, 20), class = 'b', coef = 'voicingvoiceless'),
               # weakly informative prior for all sources of variance (very diffuse)
               prior(cauchy(0,100), class = 'sd'),
               # weakly informative prior for residual variance (very diffuse)
               prior(cauchy(0,100), class = 'sigma'),
               # correlations
               prior(lkj(1), class = 'cor'))

# Run models
xmdl_10 <- brm(mdl_formula,
            family = gaussian(),
            prior = my_priors_10,
            file = "../../models/in_sesoi_10",
            data = incNeutr,
            seed = 42)

xmdl_20 <- brm(mdl_formula,
            family = gaussian(),
            prior = my_priors_20,
            file = "../../models/in_sesoi_20",
            data = incNeutr,
            seed = 42)

# Find Bayes factors for rope of 5
bayesfactor_rope(posterior = xmdl_10,
                 parameter = 'b_voicingvoiceless',
                 null = rope_narrow,
                 seed = 42) # 6.79

bayesfactor_rope(posterior = xmdl_20,
                 parameter = 'b_voicingvoiceless',
                 null = rope_narrow,
                 seed = 42) # 3.12
```

If we consider ±5ms a meaningful difference in vowel length, i.e. a ROPE of (-5, 5), specifying somewhat less diffuse priors returns Bayes factors within the realm of moderate evidence in favor of the alternative hypothesis. This makes the conclusion that this data supports, but does not prove, incomplete neutralisation seem reasonably robust.

# References

<!-- References will auto-populate in the refs div below -->

::: {#refs}
:::
