---
title: "session08 - homework"
format: html
editor: visual
date: 2025-07-17
author: Timo Roettger
execute:
  error: false
  warning: false
  message: false
  cache: true
bibliography: ../resources/bibliography.bib
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, tidybayes, brms, bayesplot, bayestestR, modelr)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

# options to increase efficiency of brms models (optional)
#options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

# Exercise

We will be re-analyzing the data from @winter2013making.

```{r}
#| label: preprocessing_sim

# Load our datasets
sim <- read_csv("../../data/linguistic_similarity.csv") |> # Changed the filepath bc this is now in hw solutions folder
  mutate(Sem = as.factor(Sem),
         Phon = as.factor(Phon))
sim


# check means
means <- sim |>
  group_by(Sem, Phon) |>
  summarise(mean_dist = mean(Distance, na.rm = TRUE))

# what's the mean of means?
mean(means$mean_dist)

```

We will now retrace the original article's complete analysis. In their original analysis, they were interested in whether (a) semantic similarity between cities, (b) phonological similarity between cities, and (c) their interaction affects how close/far apart participants draw the cities.

## (a) Specify Priors

Specify the model formula and appropriate weakly informative priors for all relevant predictors. Contrast-code the predictors.

```{r}
#| label: model_formula_sim

contrasts(sim$Sem) <- c(-0.5, 0.5)
contrasts(sim$Phon) <- c(-0.5, 0.5)

model_formula <- Distance ~ Sem * Phon

get_prior(model_formula, data = sim)
# For some reason, the coefficients for `sem` are Phon1, Sem1 etc.,
# instead of PhonSimilar, SemSimilar, etc. Why can't the machine tell that these are
# categorical values? Maybe because they're factorised this time?

my_priors <- c(prior(normal(100, 40), class = 'Intercept'), # Covers a good proportion of 297mm, the length of an A4 paper
               prior(normal(0, 40), class = 'b', coef = 'Phon1'),
               prior(normal(0, 40), class = 'b', coef = 'Sem1'),
               prior(normal(0, 40), class = 'b', coef = 'Sem1:Phon1')) # This spread for interaction is maybe a little excessive, but I don't really know how how tight it might be
# No random effects based on the data we have, yayyy


```

## (b) Run the appropiate model for the research question

```{r}
#| label: model_sim

#YOUR MODEL HERE

xmdl_sim <- brm(model_formula,
                data = sim,
                family = gaussian(),
                prior = my_priors,
                file = '../../models/08_homework.RDS',
                chains = 4,
                cores = 4, 
                iter = 2000,
                warmup = 1000,
                thin = 1,
                seed = 42)

# Shortcut to run model faster:

xmdl_sim
```

Yayyy that was nice and quick!

## (c) Posterior predictive check

Check the overall model fit with posterior predictive checks. If the model does not fit the data well, fix the problem (check `06_homework_solutions.qmd` for inspiration).

```{r}
#| label: model_fit

#YOUR CODE HERE
pp_check(xmdl_sim, ndraws = 100)

```

As with last time, the model is a bit more normal than the data, which has a right tail. I guess the right tail lines up with having a hard boundary at 0mm, since we know the data peaks below the halfway point (107mm) to the upper hard boundary.

Guess we should try making the model predict square root of distance instead of raw.

```{r}
#| label: model_adjusted

model_formula_sqrt <- sqrt(Distance) ~ Sem * Phon

# same priors but square rooted and rounded up for max conservativeness
priors_sqrt <- c(prior(normal(10, 7), class = 'Intercept'),
               prior(normal(0, 7), class = 'b', coef = 'Phon1'),
               prior(normal(0, 7), class = 'b', coef = 'Sem1'),
               prior(normal(0, 7), class = 'b', coef = 'Sem1:Phon1'))

xmdl_sim_sqrt <- brm(model_formula_sqrt,
                data = sim,
                family = gaussian(),
                prior = priors_sqrt,
                file = '../../models/08_homework_sqrt.RDS',
                chains = 4,
                cores = 4,
                iter = 2000,
                warmup = 1000,
                thin = 1,
                seed = 42)

# Posterior predictive check for square root model
pp_check(xmdl_sim_sqrt, ndraws = 100)

```

Given that the data doesn't show exactly a gaussian pattern, I think this model lines up with it as well as it can. Maybe once I understand different distributions more, I could try stuff like picking a distribution with a narrower peak and thicker tails for the `family` argument.

## (d) Assess sampling diagnostics

Check that chains are reasonably mixed and that sampling yielded a sufficient number of effective sample size. If there are issues, address them accordingly.

```{r}
#| label: sampling_diagnostics

# YOUR CODE HERE

# Assess chain mixing
mcmc_plot(xmdl_sim_sqrt, type = 'trace')

mcmc_dens_overlay(xmdl_sim_sqrt, pars = 'b_Intercept')
mcmc_dens_overlay(xmdl_sim_sqrt, pars = 'b_Phon1')
mcmc_dens_overlay(xmdl_sim_sqrt, pars = 'b_Sem1')
mcmc_dens_overlay(xmdl_sim_sqrt, pars = 'b_Sem1:Phon1')

# Check sample size
xmdl_sim_sqrt
```

Excellent caterpillars and very reasonable density patterns! Guess I won't be called out in class this time.

ESS fine.

## (e) Interpret the results

Now interpret the model relative to the research question and describe uncertainty, visualize posteriors, calculate posterior probabilities of direction, and calculate Bayes Factors for evidence for and against the posterior coefficients being point zero.

(Note: When you want to extract interaction terms like `b_Sem1:Phon1`, you have to put single quotation marks around them because R struggles with interpreting the colon)

```{r}
#| label: interpret_sim

#YOUR CODE HERE
xmdl_sim_sqrt
```

### Interpret model & describe uncertainty

The posterior mean of the main effect on distance drawn of semantic similarity is -0.69 square-root millimeters, with a 95% credible interval between -1.16 and -0.21. For the main effect of phonological similarity, it's 0.45 with a credible interval -0.05 to 0.93. And for their interaction, it's -0.19 with a credible interval -1.15 to 0.75. It seems like we're measuring pretty precisely if the 95% interval surrounds the mean with the largest estimated error being ±0.49 square rooted, or ±0.24 raw, millimeters. While the posterior credible interval is only compatible with negative slopes for semantic main effect, the other two effects are compatible with both positive and negative slopes.

### Visualise posteriors

```{r}
#| label: visualise_posteriors

spread_draws_full <- xmdl_sim_sqrt |>
  spread_draws(b_Sem1,
               b_Phon1,
               `b_Sem1:Phon1`)

sim_posteriors <- spread_draws_full |>
  pivot_longer(cols = c(b_Sem1, b_Phon1, `b_Sem1:Phon1`),
               names_to = 'Predictor',
               values_to = 'Effect') |>
  select('Predictor', 'Effect')

# Here's looking at it as a density plot, for `b_Sem1`
sim_posteriors |>
  ggplot(aes(x = Effect)) +
    stat_halfeye(alpha = 0.5,
                 aes(fill = Predictor,
                     colour = Predictor)) +
   geom_vline(xintercept = 0, lty = 'dashed') +
   labs(title = 'Posterior prediction distributions for main effects of semantic\nand phonological similarity, and their interaction,\non square-root distance drawn\n',
        x = "\nPosterior predicted effect",
        y = "Density\n") +
   theme_minimal()
```

Again this visualisation is unsatisfying given the pale & overlapping CIs, but it's something. If I had time I would plot the predictors on different horizontal planes + improve the colours.

### Posterior probabilities of direction

```{r}
#| label: posterior_pd

posteriors_horizontal <- spread_draws_full |>
  select(b_Sem1, b_Phon1, `b_Sem1:Phon1`)

posteriors_horizontal |>
  p_direction()

```

### Bayes factors

```{r}
#| bayes_factors

xmdl_sim_sqrt_priors <- brm(model_formula_sqrt,
                            data = sim,
                            family = gaussian(),
                            prior = priors_sqrt,
                            sample_prior = 'only',
                            file = '../../models/08_homework_sqrt_priors.RDS',
                            chains = 4,
                            cores = 4,
                            iter = 2000,
                            warmup = 1000,
                            thin = 1,
                            seed = 42)

priors <- xmdl_sim_sqrt_priors |>
  spread_draws(b_Sem1,
               b_Phon1,
               `b_Sem1:Phon1`) |>
  select(b_Sem1,
         b_Phon1,
         `b_Sem1:Phon1`)

bayesfactor_parameters(posteriors_horizontal, priors, direction = "two-sided", null = 0)
```

## (g) Write up your entire analysis:

The posterior mean of the main effect on distance drawn of semantic similarity is -0.69 square-root millimeters, with a 95% credible interval between -1.16 and -0.21. For the main effect of phonological similarity, it's 0.45 with a credible interval -0.05 to 0.93. And for their interaction, it's -0.19 with a credible interval -1.15 to 0.75. It seems like we're measuring pretty precisely if the 95% interval surrounds the mean with the largest estimated error being ±0.49 square rooted, or ±0.24 raw, millimeters. While the posterior credible interval is only compatible with negative slopes for semantic main effect, there are some samples over zero. The phonological main effect and interaction effect's 95% intervals are compatible with both positive and negative slopes.

Given this model, the data, and our priors, the posterior probability of semantic similarity decreasing square-root distance is \~100%; i.e. p(b\<0) = 1.0. The posterior probability of phonological similarity increasing square-root distance is \~97%; p(b\>0) = 0.97. And the posterior probability of the combination of semantic and phonological similarity decreasing square-root distance is \~65%; p(b\<0) = 0.65.

Subsequently, we calculate the Savage-Dickey density ratio [@DickeyLientz1970-The-Weighted-Li; @WagenmakersLodewyckx2010-Bayesian-hypoth; @roettgermeaningful], computing the Bayes factors against the point-null value. Our analysis yields respective Bayes factors of 2.37, 0.180, and 0.078, indicating that the probability of any of these effects being point null is actually relatively high. This suggests all evidence against the null hypothesis of these effects on square-root distance being point null [@lee-wagenmakers_Bayesian_book_2014] is anecdotal, and in some cases there is moderate evidence in favor of the null hypothesis.

## OPTIONAL (h) compare your interpretation to the original findings

YOUR COMPARISON HERE

# References

<!-- References will auto-populate in the refs div below -->

::: {#refs}
:::
